Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd
Edits to intro qualities section.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. 
<!-- In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations.  -->
Community leadership functions in WEIRD populations will likely be increasingly focused on economic contexts, creating and managing organizational structures, and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There are some broad commonalities in the qualities of community leaders across diverse populations. When strong community leaders emerge, they are most typically high status males who embody the qualities most respected by the group and necessary for successful leadership in the contexts in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 (and others) in conceptualizing social status as a relative indicator of the social value of the individual by the group and of their relative access to contested resources. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014], however, if there is a causal relationship it is likely complex, context-specific, and driven by underlying correlates.

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. @yanca_female_2004 found that sex-differences in political influence was more likely driven by ecological and social factors, rather than sex (or gender identity) per se, including social networks and access to same-sex allies. Two recent anthropological studies relying on ethnographic fieldwork among small-scale societies and comparing leaders and non-leaders, revealed sex is not a strong predictor of community influence or elected leadership, after accounting for economic specializations and a broad range of demographic, phenotypic, cognitive, and social traits [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Across cultural contexts elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers in businesses across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

<!-- NON WEIRD QUALITIES -->
Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012]. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement [@grabo_charismatic_2016].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. In such contexts community leaders often lack strong coercive authority, but maintain influence through ideological systems and ritual or religious statuses inaccessible to non-elites [@stanish_evolution_2010]. Within state-level societies including nonindustrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

<!-- INTELLIGENCE -->
Across disciplines many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009]. Recent cross-cultural evidence and from a small-scale society implicates strong relationships between perceptions of intelligence and leadership among males [@garfield_evolutionary_2019; @garfield_investigating_2019]. These results suggests high intelligence and specialized knowledge are universal traits of political and community leaders. Within managerial studies scholars often suggest the cognitive demands of leadership escalate with increasing organizational complexity [@zaccaro_organizational_2002]. More complex organizations require leaders equipped to process complex information, understand distinct departmentalized bodies, and resolve conflicting points of view, as well as be equipped to socially manage diverse individuals [@bass_handbook_1990; @schneider_organizations_2006; @antonakis_does_2009]. However, recent evidence suggests a curvilinear effect in that exceptionally intelligence individuals at the high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

<!--  CONCLUSION PARAGRAPH/STATEMENT. -->
Much of the theoretical leadership literature across disciplines suggests relative uniformity in many qualities of community leaders. Intelligence, charisma, effective decision-making capabilities, high social status and strong social networks are likely to be universal qualities of political and community leaders. In more socially stratified, politically complex, and in increasingly WEIRD contexts, heredity, ascribed social statuses and affiliations are more likely to be critical qualities in achieving and/or maintaining leadership positions. We also envision a curvilinear relationship between leadership status and supernatural abilities and qualities. Community leaders among small-scale hunter-gatherers may be seen as experts in the use of ritual and supernatural "tools" though they typically do not have a monopoly in their use. In chiefdoms and and in early prehistoric state societies political leaders are often seen as divine and have special supernatural qualities commoners do not. In contemporary WEIRD societies there is a trend of secularization among political leaders, generally, although religious and spiritual leaders do have substantial influence over social and moral norms.

>>>>

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. 
<!-- The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies.  -->
In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. 
<!-- In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations.  -->
Community leadership in WEIRD populations is likely be largely focused on economic functions, creating and managing organizational structures, and managing inter-group relations, whereas in less WEIRD contexts community leadership functions are more likely to focus on resolving interpersonal conflicts and managing intra-group dynamics.

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There are some broad commonalities suggested in the qualities of community leaders across diverse populations. When strong community leaders emerge, they are most likely to be high status males who embody the qualities most respected by the group and necessary for successful leadership in the contexts in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 (and others) in conceptualizing social status as a relative indicator of the social value of the individual by the group and of their relative access to contested resources. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014], however, any causality between the two is likely complex, context-specific, and driven by underlying correlates.

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. @yanca_female_2004 found that sex-differences in political influence was more likely driven by ecological and social factors, such as social networks and access to same-sex allies, rather than sex (or gender identity) per se. Two recent studies relying on ethnographic fieldwork comparing leaders and non-leaders in small-scale societies, both revealed sex is not a strong predictor of community influence or elected leadership, after accounting for economic specializations and a broad range of demographic, phenotypic, and social traits [@garfield_investigating_2019; @von_rueden_sex_2018].

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Across cultures elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers in businesses across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

<!-- NON WEIRD QUALITIES -->
Among small-scale, autonomous, relatively egalitarian societies community leaders tend to be older, respected males, with large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012]. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders, but thought to be especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement [@grabo_charismatic_2016].

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. In such contexts community leaders often lack strong coercive authority, but maintain influence through ideological systems and ritual or religious statuses inaccessible to non-elites [@stanish_evolution_2010]. Within state-level societies including nonindustrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].

<!-- INTELLIGENCE -->
Many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], including across diverse societies [@garfield_evolutionary_2019; @garfield_investigating_2019]. These results suggests high intelligence and specialized knowledge are potentially universal traits of political and community leaders. Within managerial studies scholars often suggest the cognitive demands of leadership escalate with increasing organizational complexity [@zaccaro_organizational_2002]. More complex organizations are thought to require leaders equipped to process complex information, understand distinct departmentalized bodies, and resolve conflicting points of view among diverse individuals [@bass_handbook_1990; @schneider_organizations_2006; @antonakis_does_2009]. Some recent evidence, however, suggests a curvilinear effect in that exceptionally intelligence individuals at the high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

<!--  CONCLUSION PARAGRAPH/STATEMENT. -->
Much of the theoretical leadership literature across disciplines suggests relative uniformity in many qualities of community leaders. Intelligence, charisma, effective decision-making capabilities, high social status and strong social networks are likely to be broadly representative qualities of political and community leaders. In more socially stratified, politically complex, and in increasingly WEIRD contexts, heredity, ascribed social statuses and affiliations are more likely to be critical qualities in achieving and/or maintaining leadership positions.

<!-- We also envision a curvilinear relationship between leadership status and supernatural abilities and qualities. Community leaders among small-scale hunter-gatherers may be seen as experts in the use of ritual and supernatural "tools" though they typically do not have a monopoly in their use. In chiefdoms and and in early prehistoric state societies political leaders are often seen as divine and have special supernatural qualities commoners do not. In contemporary WEIRD societies there is a trend of secularization among political leaders, generally, although religious and spiritual leaders do have substantial influence over social and moral norms.    -->
